<link href="http://kevinburke.bitbucket.org/markdowncss/markdown.css" rel="stylesheet"></link>
<link href="http://kbroman.org/qtlcharts/assets/vignettes/markdown_modified.css" rel="stylesheet"></link>

## Knitr/R Markdown example

This is a simple example using knitr and R markdown to mix code and
text.

We'll start by setting the seed for the random number generator.

```{r set_seed}
set.seed(53079239)
```

We'll first simulate some data with $x \sim N(\mu=10, \sigma=5)$ and
$y = 2x + \epsilon$, where $\epsilon \sim N(\mu=0, \sigma=2)$.

```{r simulate}
x <- rnorm(100, 10, 5)
y <- 2*x + rnorm(100, 0, 2)
```

Here's a scatterplot of the data.

```{r scatterplot}
plot(x, y, pch=21, bg="slateblue", las=1)
```

The estimated correlation between $x$ and $y$ is
`r sprintf("%.2f", cor(x,y))`.

Here is the result of regressing $y$ on $x$.

```{r regression}
out <- lm(y ~ x)
summary(out)
```

So note that the estimated slope for the regression line is
`r sprintf("%.2f", out$coef[2])`.

### Tables

To include a table, use `kable` in the knitr package or `xtable` in
the xtable package.

`kable` is simpler but has few options.

```{r kable, results="asis"}
coef_tab <- summary(out)$coef
kable(summary(out)$coef, format="html",
      digits=2)
```

`xtable` gives you more complete control.

```{r xtable, results="asis"}
library(xtable)
tab <- xtable(coef_tab, digits=c(0, 2, 2, 1, 3))
print(tab, type="html")
```

### It's usually best to end with session information

```{r sessionInfo}
sessionInfo()
```
