---
title: "knitr/R Markdown example"
author: "Karl Broman"
date: "28 January 2015"
output: html_document
---

This is a simple example using knitr and R markdown to mix code and
text.

We'll start by setting the seed for the random number generator.

```{r set_seed}
set.seed(53079239)
```

We'll first simulate some data with $x \sim N(\mu=10, \sigma=5)$ and
$y = 2x + \epsilon$, where $\epsilon \sim N(\mu=0, \sigma=2)$.

```{r simulate}
x <- rnorm(100, 10, 5)
y <- 2*x + rnorm(100, 0, 2)
```

Here's a scatterplot of the data.

```{r scatterplot}
plot(x, y, pch=21, bg="slateblue", las=1)
```

The estimated correlation between $x$ and $y$ is
`r sprintf("%.2f", cor(x,y))`.

Here is the result of regressing $y$ on $x$.

```{r regression}
out <- lm(y ~ x)
summary(out)
```

So note that the estimated slope for the regression line is
`r sprintf("%.2f", out$coef[2])`.

### Tables

To include a table, I recommend the
[pander package](http://rapporter.github.io/pander/#markdown-tables).
Use `panderOptions()` to change various options, and `pander()` to
create the actual table. If you call `pander` with the output of
`lm()`, the default is to give the table of regression coefficients.

```{r pander_table}
library(pander)
panderOptions('digits', 3)
pander(out, caption="Regression coefficients")
```

### It's usually best to end with session information

```{r sessionInfo}
sessionInfo()
```
